---
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr);library(knitr); library(RColorBrewer)
library(ggplot2);library(gridExtra)
library(scales); library(foreach)
library(plotly); library(grid)

'%ni%' <- Negate('%in%')

opts_chunk$set(warning=FALSE, message=FALSE, echo=FALSE, row.names = FALSE, cache=FALSE,
               fig.align = 'center',  fig.width=6, fig.height=4, results = 'hide')
options(knitr.kable.NA = '') 

source("C:/Dropbox/Professional/Grants_Current/SS Chemistry Data Mining/code/helper functions.R")
raw <-read.delim("C:/Dropbox/Professional/Grants_Current/SS Chemistry Data Mining/data/chem_data_20170227.txt", sep = "\t")

ew <- raw %>% filter(((term_idx == "F14" & section==1) | crt==1)  &
                        !is.na(gender) & !is.na(stulevel) & grade %ni% c('W ', 'I ')) %>% 
      droplevels() %>%
      mutate(repeat_grade = ifelse(grade %in% c('D ', 'D+', 'F ', 'WU'), 1, 0),
             pass = factor(repeat_grade, labels=c("Pass", "Fail")),
             term_idx = factor(term_idx, levels=c("F14","F15", "S16", "F16")),
             DSS = ifelse(dss.1==1 | dss == 'Y', 'DSS', 'not in DSS'), 
             soc = factor(soc, levels=c('Non-SOC', 'SOC', 'Multiple/Mixed')), 
             eop = factor(eop, labels=c("Non-EOP", "EOP"))) %>%
      select(-dss.1, -dss, -erss_stulevel, erss_enrstat)

ew$crt[ew$term_idx == "F16"] <- 2
ew$crt <- factor(ew$crt, labels=c("Traditional", "CRT", "CRT+SI"))

class <- ew %>% group_by(year, term, term_idx, section, crt) %>%
          summarise(classgpa = mean(gpa), dfw = sum(repeat_grade), 
                    N = n(), dfw_rate = dfw/N)

f146 <- filter(ew, term_idx %in% c("F14", "F16")) %>% droplevels()

# viz opts
yearlab <- c(2009, rep(2010:2016, each=2))
term.col <- brewer.pal(9, "Set1")[c(7, 3)]
crt.col <- brewer.pal(9, "Set1")[1:3]
```

# Analysis Overview

To analyze the effect of the course redesign, we consider data from Fall 2014 to Fall 2016, for all sections of Chemistry 111. Students withdrawing or receving an incomplete (`W/I`) are exluded from this analysis. A repeatable grade is defined as a student reciving a *D*, *D+*, *F* grade, or an unexcused withdrawl *WU*. 

1. Examine bivariate associations between outcome measures, covariates and the CRT intervention condition. 
    a. Student and class level characteristics are used as covariates. 
    b. Each predictor is compared to the intervention condition (CRT vs Traditional), and to the outcomes (GPA and receipt of a repeatable grade.)
    c. Analysis is done on two sets of data samples: 
        - All terms (Full sample) 
        - Just F14 and F16 terms only 
2. Model building - Start with variables that were shown to be bivariately associated with either intervention or outcome, build a multivariable statistical model to assess the effect of the CRT on student performance. 
3. Use a supervised machine learning technique to identify other key characteristics highly predictive of a student recieving a repeatable grade. 


### Outcome measurements

**1. GPA**

Since GPA is being treated as a continuous variable, distributions of GPA are displayed using boxplots with overlaid violin plots. The diamonds represent the average or mean value for that group. Two-sample T-tests and ANOVA tests are used to compare the average GPA across levels of the covariate of interest. 

**2. Receipt of a repeatable grade**

We also dichotomize the grade a student received and create an indicator for whether or not the student received a repeatable grade. This is also called the _failure rate_ in some places in this report. 

Where the sample size is large enough, $\chi^{2}$ tests for equal proportions across two / and multiple groups are conducted, otherwise Fishers Exact Tests are used. Any comparison where a zero cell occurs (a combination of factors that does not exist in the data), no statistical test is conducted. 

# Performance Outcomes

We examine four semesters of data from a single instructor. One semester of traditional course offering from F14, two semesters of redesigned courses (F15, S16) and one semester of the redesign with an SI enhancement. These are colored in red, blue and green respectively in most plots. 

### GPA
The overall distribution of gpa for the crt courses has shifted in the positive direction compared to the traditional courses. The mean for Fall 15 and F16 are higher than F14, with S16 closer to F14 than the other redesigned terms. The variance in GPA from the last semester is much higher than the rest, potentially indicating that the effect of SI did not have a uniform impact on increasing the GPA for all students. 

```{r}
ggplot(ew, aes(y=gpa, x=term_idx, fill=crt)) + 
        geom_boxplot(alpha=.5, width=.3) +
        geom_violin(alpha=.3) + xlab("") + theme_bw() + ylab("GPA") + xlab("Semester") + 
        scale_fill_manual(values=crt.col, guide=FALSE) + 
        stat_summary(fun.y=mean, geom="point", shape=18, size=4)
```


### DWF rate
The percent of students receiving a repeatable grade has declined over the past four semesters. 
```{r}
ggplot(class, aes(y=dfw_rate, x=term_idx, col = crt, shape=crt, label=paste0(round(dfw_rate*100,1),"%"))) + 
          geom_text(nudge_y=.05)  + geom_point(size=5) + ylab("") + theme_bw() + 
          scale_color_manual(values=crt.col, name="Class Type") + scale_shape_discrete(guide=FALSE) +  
          scale_y_continuous(limits=c(0,1), labels=scales::percent) + xlab("") 
```


### Statistical Comparisons
```{r, results='show'}
cl <- class %>% ungroup() %>% arrange(term_idx) %>% select(term_idx, crt, classgpa, dfw, N, dfw_rate)
names(cl) <- c("Term", "Type", "Ave. GPA", "# DFW", "Enrollment", "DFW Rate")
kable(cl,
      digits=2, align='c', row.names=FALSE, caption='Class level performance characteristics')
```

```{r}
gpa.crt.ew <- univ.gpa(ew, col.var = 'crt')
crt.p <- format.pval(unlist(summary(aov(gpa.crt.ew$test)))[9], digits=2)

gpa.crt.f146 <- univ.gpa(f146, col.var = 'crt')
```

* There is a significant improvement in GPA averaged across all redesigned sections compared to traditional (p=`r crt.p`). 
* There is also a significant improvement in average GPA in F16 compared to F14 (p=`r format.pval(gpa.crt.f146$test$p.value, digits=2)`). 
* The failure rate in F16 is lower than the failure rate in F14, but only marginally so (p=`r format.pval(chisq.test(ew$repeat_grade, ew$crt)$p.value, digits=2)`)
* The proportion of repeatable grades does not differ significantly across semesters (p=`r format.pval(chisq.test(f146$repeat_grade, f146$crt)$p.value, digits=2)`)


# Student Characteristics 

## Gender
#### Course Type

The gender distribution remained constant across the past four semesters.

```{r}
g.term <- biv.n.row.pct(df=ew, row.var = 'gender', col.var = 'term_idx')
chisq.test(table(ew$gender, ew$term_idx))
t.test((gender=="F") ~ term_idx, data=f146)
```

```{r, results='show'}
kable(g.term$table, row.names = FALSE)

g.term$plot + scale_y_continuous(labels=scales::percent, limits=c(.25, .75)) + theme_bw()
```


#### Performance - GPA
There is no statistical difference in the average GPA between males and females, regardless of cohort. 

```{r}
gpa.gender.ew   <- univ.gpa(ew, col.var = 'gender')
gpa.gender.f146 <- univ.gpa(f146, col.var = 'gender')

a <- data.frame(gpa.gender.ew$table, 
               p = c(format.pval(gpa.gender.ew$test$p.value, digits=2), NA), 
               gpa.gender.f146$table[,-1],
               p2 = c(format.pval(gpa.gender.f146$test$p.value, digits=2), NA)
               )

names(a) <- c("Gender", "Full mean", "Full SD", "p-value",  "F14 & F16 mean", "F14 & F16 SD", "p-value")
```

```{r, results='show'}
kable(a, digits=2, align='c')
```

```{r, fig.width=12}
a <- gpa.gender.ew$plot + ggtitle("Full Sample") + xlab("")
b <- gpa.gender.f146$plot + ggtitle("F14 & F16 only") + xlab("")

grid_arrange_shared_legend(a, b, ncol=2, nrow=1, position="right")
```


#### Performance - DFW rate
There is no difference in proportion of males and females getting a repeatable grade on the entire sample, or when using data on the F14 and F16 cohorts only.

```{r}
gender.dwf.ew <- biv.n.row.pct(ew, col.var = 'gender', row.var='repeat_grade')
gender.dwf.f146<- biv.n.row.pct(f146, col.var = 'gender',row.var='repeat_grade')
```

```{r, results='show'}
a <- data.frame(g = c("No", "Yes"),
           gender.dwf.ew$table[,-1], 
           p=c(format.pval(chisq.test(ew$repeat_grade, ew$gender)$p.value, digits=2), NA), 
           gender.dwf.f146$table[,-1], 
           p=c(format.pval(chisq.test(f146$repeat_grade, f146$gender)$p.value, digits=2), NA)
           )
names(a) <- names(a) <- c("Repeatable Grade", "N (%) Female \n Full Sample", "N(%) Male \n Full Sample", "p-value",  
                          "N(%) Female \n F 14 & 16", "N(%) Male \n F14 & F16", "p-value")

kable(a, row.names = FALSE, align='c')
```

## Age 

#### Course Type

Students in F14 were on average 0.68 (0.01, 1.34) years older than those in the two CRT semesters (p=.04), and 0.87 (0.1, 1.65) years older than students in Fall 16 (p=.02).

```{r}
age.crt.ew   <- biv.mean(df=ew, var='age', group='crt')
age.crt.ew$plot + scale_fill_discrete(name="Class Type")
```

```{r, results='show'}
kable(biv.mean.wide(df=ew, var='age', group='crt'), align='c', caption="Mean and SD for ages within class types")
```

```{r, results='show'}
kable(TukeyHSD(aov(age~crt, data=ew))$crt, digits=2, 
      caption="Tukeys Honest Significant Difference for differences between all pairs")
```


#### Performance - GPA
GPA and Age have a non-linear relationship. For students less than 25 years of age, the pattern of relationship between age and GPA does not seem to differ greatly across the three course types. This indicates that the intervention has a similar affect on a students GPA regardless of their age. 
```{r, fig.width=10}
a <- ggplot(ew, aes(x=age, y=gpa, col=crt)) + 
        geom_point(alpha=.5) + geom_smooth(se=FALSE) + theme_minimal() + 
        scale_color_manual(values=crt.col)  + ggtitle("Everyone")

b <- ggplot(filter(ew, age<25), aes(x=age, y=gpa, col=crt)) + 
        geom_point(alpha=.5) + geom_smooth(se=FALSE) + theme_minimal() + 
        scale_color_manual(values=crt.col) + ggtitle("Age < 25yrs")

grid_arrange_shared_legend(a, b, ncol=2, nrow=1, position="bottom")
```

```{r}
out <- mlr.table(lm(gpa~age + crt, data=ew))
out
```

#### Performance - DWF rate [[NOT DONE YET]]



## Race / Ethnicity
The demographic shift across all chem classes is displayed in the Exploratory Analysis document. How does this compare to the campus-wide shift? 

```{r}
eth.term <- biv.n.row.pct(df=ew, row.var = 'race_code', col.var = 'term_idx')
kable(eth.term$table, row.names = FALSE)
eth.term$plot
```

There are too many ethnicities with low sample sizes to compare statistically. Here we just look at two smaller groupings of race/ethnicity. 


## Underrepresented Minority (URM)
* Non-URM: Asian, NHOPI, White
* URM: Black, HL
* UNK: Mult, Unk

#### CRT vs. Traditional 

**Full Sample**
The distribution of URM across the past four terms has changed significantly. 
```{r}
urm.term <- biv.n.row.pct(df=ew, row.var = 'urm_status', col.var = 'term_idx')
kable(urm.term$table, row.names = FALSE)
urm.term$plot
chisq.test(table(ew$urm_status, ew$term_idx))
```

**F14 & F16 only**
No statistical difference in the distribution of proportions between years. 
```{r}
urm.146 <- biv.n.row.pct(df=f146, row.var = 'urm_status', col.var = 'term_idx')
kable(urm.146$table, row.names = FALSE)
chisq.test(table(f146$urm_status, f146$term_idx))
```


#### Performance - GPA
**Full Sample**
URM students have significantly lower GPA (.38 lower) compared to non-URM students. The GPA for students with unknown URM status have marginally lower (.23) GPA compared to non-URM students. 

```{r}
gpa.urm.ew <- univ.gpa(ew, col.var = 'urm_status')
kable(gpa.urm.ew$table, digits=2)
gpa.urm.ew$plot
summary(gpa.urm.ew$test)
```

**F14 & F16 only**
URM students have marginally lower GPA (.23 lower) on average compared to non-URM students. The GPA for students with unknown URM status is not different than the average GPA for non-URM students. 

```{r}
gpa.urm.f146 <- univ.gpa(f146, col.var = 'urm_status')
kable(gpa.urm.f146$table, digits=2)
gpa.urm.f146$plot
summary(gpa.urm.f146$test)
```

#### Performance - DWF rate

**Full Sample**
There is a statistical difference in DWF rate by URM status in the full cohort.
```{r}
urm.dwf.ew <- biv.n.row.pct(ew, col.var = 'urm_status', row.var='repeat_grade')
kable(urm.dwf.ew$table, row.names = FALSE)
chisq.test(ew$repeat_grade, ew$urm_status)
```

**F14 & F16 only**
There is no difference in dwf rate by URM status comparing F14 to F16 only. 
```{r}
urm.dwf.f146 <- biv.n.row.pct(f146, col.var = 'urm_status', row.var='repeat_grade')
kable(urm.dwf.f146$table, row.names = FALSE)
chisq.test(f146$repeat_grade, f146$urm_status)
```


## Students of Color (SOC)
* Non-SOC= White
* SOC = AIAN, Asian, Black, HL, NHOPI
* Multiple/Mixed = Multi, Unk


#### CRT vs. Traditional 

**Full Sample**
The distribution of SOC across the past four terms has changed significantly (p=.01)
```{r}
soc.term <- biv.n.row.pct(df=ew, row.var = 'soc', col.var = 'term_idx')
kable(soc.term$table, row.names = FALSE)
soc.term$plot
chisq.test(table(ew$soc, ew$term_idx))
```

**F14 & F16 only**
No statistical difference in the distribution of proportions between F14 & F16. 
```{r}
soc.146 <- biv.n.row.pct(df=f146, row.var = 'soc', col.var = 'term_idx')
kable(soc.146$table, row.names = FALSE)
chisq.test(table(f146$soc, f146$term_idx))
```


#### Performance - GPA
**Full Sample**

Non-SOC students have a significantly higher GPA compared to students of color (.39 higher),and students identifying as Multiple/Mixed ethnities (.27 higher). 
```{r}
gpa.soc.ew <- univ.gpa(ew, col.var = 'soc')
kable(gpa.soc.ew$table, digits=2)
gpa.soc.ew$plot
summary(gpa.soc.ew$test)
```

**F14 & F16 only**
Only comparing F14 and F16, non-SOC students have a significantly higher GPA compared to students of color (.28 higher), and marginally higher than those from Multiple/Mixed ethnicities (.29 higher). 

```{r}
gpa.soc.f146 <- univ.gpa(f146, col.var = 'soc')
kable(gpa.soc.f146$table, digits=2)
gpa.soc.f146$plot
summary(gpa.soc.f146$test)
```

#### Performance - DWF rate

**Full Sample**
There is strong statistical difference in DWF rate by SOC status in the full cohort.
```{r}
soc.dwf.ew <- biv.n.row.pct(ew, col.var = 'soc', row.var='repeat_grade')
kable(soc.dwf.ew$table, row.names = FALSE)
chisq.test(ew$repeat_grade, ew$soc)
```

**F14 & F16 only**
There is a marginal difference in dwf rate by SOC status when looking at F14 and F16 only. 
```{r}
soc.dwf.f146 <- biv.n.row.pct(f146, col.var = 'soc', row.var='repeat_grade')
kable(soc.dwf.f146$table, row.names = FALSE)
chisq.test(f146$repeat_grade, f146$soc)
```


# Student Groups 

## EOP
The number of EOP students is too small to make accurate comparisons. 

#### CRT vs. Traditional 

**Full Sample**
The distribution of EOP student status has not changed significantly. 
```{r}
eop.term <- biv.n.row.pct(df=ew, row.var = 'eop', col.var = 'term_idx')
kable(eop.term$table, row.names = FALSE)
chisq.test(table(ew$eop, ew$term_idx))
```

#### Performance - GPA
**Full Sample**
There is no significant difference in average GPA for students in EOP compared to non-EOP students. 
```{r}
gpa.eop.ew <- univ.gpa(ew, col.var = 'eop')
kable(gpa.eop.ew$table, digits=2)
gpa.eop.ew$plot
gpa.eop.ew$test
```

**F14 & F16 only**
EOP students did significantly better compared to non-EOP students when looking at the F14 and F16 cohorts only. 

```{r}
gpa.eop.f146 <- univ.gpa(f146, col.var = 'eop')
kable(gpa.eop.f146$table, digits=2)
gpa.eop.f146$plot
gpa.eop.f146$test
```

#### Performance - DWF rate

**Full Sample**
There is no difference in the % of repeatable grades between EOP and non-EOP students. 
```{r}
eop.dwf.ew <- biv.n.row.pct(ew, col.var = 'eop', row.var='repeat_grade')
kable(eop.dwf.ew$table, row.names = FALSE)
chisq.test(ew$repeat_grade, ew$eop)
```

**F14 & F16 only**
There is no difference in the % of repeatable grades between EOP and non-EOP students.
```{r}
eop.dwf.f146 <- biv.n.row.pct(f146, col.var = 'eop', row.var='repeat_grade')
kable(eop.dwf.f146$table, row.names = FALSE)
fisher.test(f146$repeat_grade, f146$eop)
```


## Any formal student group (FSG)



## First Generation

#### CRT vs. Traditional 

**Full Sample**
The distribution of first generation student status has not changed significantly across the past four terms. 
```{r}
firstgeneration.term <- biv.n.row.pct(df=ew, row.var = 'firstgeneration', col.var = 'term_idx')
kable(firstgeneration.term$table, row.names = FALSE)
chisq.test(table(ew$firstgeneration, ew$term_idx))
```

#### Performance - GPA
**Full Sample**
First generation students have significantly lower GPA compared to non-first gen students. 
```{r}
gpa.firstgeneration.ew <- univ.gpa(ew, col.var = 'firstgeneration')
kable(gpa.firstgeneration.ew$table, digits=2)
gpa.firstgeneration.ew$plot
gpa.firstgeneration.ew$test
```

**F14 & F16 only**
When comparing the F14 and F16 cohorts only, this difference disappears. 

```{r}
gpa.firstgeneration.f146 <- univ.gpa(f146, col.var = 'firstgeneration')
kable(gpa.firstgeneration.f146$table, digits=2)
gpa.firstgeneration.f146$plot
gpa.firstgeneration.f146$test
```

#### Performance - DWF rate

**Full Sample**
A significantly higher proportion of first generation students received repeatable grades. 
```{r}
firstgeneration.dwf.ew <- biv.n.row.pct(ew, col.var = 'firstgeneration', row.var='repeat_grade')
kable(firstgeneration.dwf.ew$table, row.names = FALSE)
chisq.test(ew$repeat_grade, ew$firstgeneration)
```

**F14 & F16 only**
Again, this difference disappears when considering only F14 vs F16 cohorts. 
```{r}
firstgeneration.dwf.f146 <- biv.n.row.pct(f146, col.var = 'firstgeneration', row.var='repeat_grade')
kable(firstgeneration.dwf.f146$table, row.names = FALSE)
fisher.test(f146$repeat_grade, f146$firstgeneration)
```



# College Prep
This section examines the relationship between the amount and type of college prep
courses (HS and college) a student has at the time of admission, and the student's 
gpa in Chem 111.


## Admissions Index (HS Eligibility)

```{r}
ggplot(ew, aes(y=admission_index, x=term_idx, fill=crt)) + geom_boxplot(alpha=.5, width=.4) +
       geom_violin(alpha=.3) + xlab("") + 
       scale_fill_manual(values=crt.col, guide=FALSE) + 
       stat_summary(fun.y=mean, geom="point", shape=18, size=4) 
```

Average admission index is marginally statistically different across years. 
```{r}
summary(aov(admission_index ~ term_idx, data=ew))
```

#### GPA
In the following scatterplots, each dot represents one student. Due to the discrete nature of gpa and number of classes, many points are overlaid on the same spot. So the darker the point, the more students are represented there. 

```{r}
ggplot(ew, aes(x=admission_index, y=gpa, col=crt)) + 
        geom_point(alpha=.5) + geom_smooth(se=FALSE) + theme_minimal() + 
        ggtitle("GPA by Admission index - by CRT group") + 
        scale_color_manual(values=crt.col) 

ggplot(f146, aes(x=admission_index, y=gpa, col=crt)) + 
        geom_point(alpha=.5) + geom_smooth(se=FALSE) + theme_minimal() + 
        ggtitle("GPA by Admission index - for Fall '14 and '16 only") + 
        scale_color_manual(values=crt.col) 
```

Admissions index does not have a strictly linear relationship with GPA. 
A standard linear model is not appropriate, a spline model with a knot around 3.5 should be used instead. <span style="color:red">Do this</span>

#### DFW 
```{r}
ggplot(ew, aes(y=admission_index, x=pass, fill=pass)) +  
       geom_boxplot(alpha=.5, width=.4) +
       geom_violin(alpha=.3) + xlab("") + 
       scale_fill_manual(values=crt.col, guide=FALSE) + 
       stat_summary(fun.y=mean, geom="point", shape=18, size=4) + 
       theme_minimal() + 
       ggtitle("Distribution of admissions index by success status.") 
```

There is a strong relationship between admissions index and whether or not
the student will pass the class. 
```{r}
t.test(admission_index ~ pass, data=ew)
```

## ELM Status


## EPT Status


## GE Status




# Bivariate Summary

Factors that were associated with GPA or passing the class

* Admissions index - but not linearly
* 

Comparing the entire cohort to F14 and F16 only -- often differences seen in the entire cohort were not seen when only considering the F14 and F16 cohorts only. Even though there is only one spring term included, this may be an indication of a "semester" effect that we have seen evidence of in other places. 


# Multivariate modeling

## GPA - recap of relationship
```{r}
ggplot(ew, aes(x=age, y=gpa, col=crt)) + 
        geom_point(alpha=.5) + geom_smooth(se=FALSE) + theme_minimal() + 
        ggtitle("GPA by age index - full sample") + 
        geom_smooth(method=lm, se=FALSE)+ 
        scale_color_manual(values=crt.col) +     
        scale_x_continuous(limits=c(17,25))
```
For those under 25, after accounting for a **linear** effect of age, the CRT significantly improves GPA. 
```{r}
summary(lm(gpa ~ age + crt, data=filter(ew, age<=25)))
```


## HS eligibility [In progress]
```{r}
ggplot(ew, aes(x=admission_index, y=gpa, col=crt)) + 
        geom_point(alpha=.5) + geom_smooth(se=FALSE) + 
        theme_minimal() + 
        ggtitle("GPA by Admission index - by CRT group") + 
        scale_color_manual(values=crt.col) 
```        

Fit a linear spline model, with a knot at 3.75. That is, 
two linear models with differing slopes.

```{r}
ew$lowai <- (ew$admission_index - 3.75)
ew$lowai[ew$admission_index < 3.75] <- 0

fit.spline <- lm(gpa~admission_index + lowai, data=ew)
plot(fit.spline$model$gpa~fit.spline$model$admission_index)
lines(sort(predict(fit.spline)) ~ sort(fit.spline$model$admission_index), lwd=2)
```

Interaction between admissions index and CRT -- allow there to be separate slopes between CRT groups in both high and low admission index areas. 
-- in other words-- does the CRT group _change_ the relationship between admissions index and GPA? 
```{r}
summary(lm(gpa~admission_index + lowai + gpa, data=ew))
summary(lm(gpa~admission_index + lowai + crt + 
             crt*admission_index + lowai*crt, data=ew))
```


---- 
Notes

```{r}
a <- lm(gpa ~ crt + gender + firstgeneration, data=ew)
summary(a)
```
* can we show that SI is the cause of the reduction of the GAP between first gen and/or URM students. 

* DFW rate as a function of attendance
* HS eligibility vs. SI visits
* 

```{r}
a <- lm(gpa ~ crt + gender + firstgeneration, data=filter(ew, term_idx %in% c('F14', 'F16')))
summary(a)
```

* add logistic regression of repeatable grade vs crt
```{r}
b <- glm(repeat_grade ~ crt + admission_index, 
         data=ew, family='binomial')
summary(b)
exp(coef(b))
```

* don't forget to add difference between chem performance and other classes 
* plot gap in gpa and dwf across term


# CART - Predicting a failing grade. 
Work in progress...

```{r, eval=FALSE}
library(rpart)
a <- rpart(repeat_grade ~ crt + age + gender, 
           data=ew, method="class")
#a <- prune(a, cp=.014)

plot(a, uniform=TRUE)
text(a, use.n=TRUE, all=TRUE, cex=.8)
```
