---
title: 'Historical Trends in Students taking Chem 111'
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

This section analyses the trends in characteristics of the student group who is taking Chemistry 111, and compares them to the campus level averages. 

We consider data from Fall 2009 - Fall 2016, for all sections of Chemistry 111. Students withdrawing or receiving an incomplete (`W/I`) are excluded from this analysis. A repeatable grade is defined as a student receiving a `D, D+`, or `F` grades, or an unexcused withdrawal `WU`. 


```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr);library(knitr); library(kableExtra); 
library(RColorBrewer); library(ggplot2);
library(gridExtra); library(grid)
library(scales); library(foreach); library(plotly)

'%ni%' <- Negate('%in%')

opts_chunk$set(warning=FALSE, message=FALSE, echo=FALSE, row.names = FALSE, cache=FALSE,
               fig.align = 'center',  fig.width=8, fig.height=5)
options(knitr.kable.NA = '') 

source("C:/Dropbox/Professional/Grants_Current/SS Chemistry Data Mining/code/helper functions.R")
rawtrends <-read.delim("C:/Dropbox/Professional/Grants_Current/SS Chemistry Data Mining/data/chem_data_20170227.txt", sep = "\t")
campus_aves <- read.delim("C:/Dropbox/Professional/Grants_Current/SS Chemistry Data Mining/data/campus_aves.txt", sep="\t")
campus_props <- read.delim("C:/Dropbox/Professional/Grants_Current/SS Chemistry Data Mining/data/campus_props.txt", sep="\t")


raw <- rawtrends %>% filter(grade %ni% c('W ', 'I ') & erss_stulevel < 5) %>% droplevels() %>%
            mutate(term_idx = factor(term_idx, 
                      levels = c("F09", "S10", "F10", "S11", "F11", "S12", "F12", "S13", 
                                "F13", "S14", "F14", "S15", "F15", "S16", "F16")),
                    stulevel = factor(erss_stulevel, labels = c("Freshman", "Sophomore", "Junior", "Senior")), 
                    DSS = ifelse(dss.1==1 | dss == 'Y', 'DSS', 'not in DSS'), 
                    repeat_grade = ifelse(grade %in% c('D ', 'D+', 'F ', 'WU'), 1, 0)) %>%
            select(-dss.1, -dss, -erss_stulevel, erss_enrstat, -dfw_rate)
                
# group number of CP classes
chem <- raw %>% mutate(cpmath = cut(cp_math, c(0,6,8,10,17), right=FALSE),
                    cplab     = cut(cp_lab, c(0,4,6,8,15), right=FALSE),
                    cpengl    = (cp_engl>0))

## Recoding d2$college_name into d2$college_name_rec
chem$college_name_rec <- as.character(chem$college_name)
chem$college_name_rec[chem$college_name == "AGR"] <- "NS"
chem$college_name_rec[chem$college_name %in%  c("BSS", "BUS", "CME", "HFA", "UGE")] <- "OTH"
chem$college_name_rec <- factor(chem$college_name_rec, levels=c("NS", "ECC", "OTH"))


# redo class level measures
class <- chem %>% group_by(year, term, term_idx, section) %>%
          summarise(classgpa = mean(gpa), dfw = sum(repeat_grade), 
                    enrollment = n(), dfw_rate = dfw/enrollment)
         
yearlab <- c(2009, rep(2010:2016, each=2))
term.col <- brewer.pal(9, "Set1")[c(7, 3)]
```


# Course Characteristics

## Enrollment
```{r}
ggplot(chem, aes(x=term_idx, y=..count.., fill=term)) + geom_bar() +
  geom_text(stat="count", aes(y = ..count.., vjust=-.3, label=..count..)) +
  xlab("Year") + ylab("Enrollment") + theme_bw() +  
  scale_x_discrete(labels=yearlab) + scale_fill_manual(values=term.col, name = "Term")
```
  
Spring term demand has steadily increased since 2009. 
 
## Class size

Fall class sizes have remained relatively constant between 140-160 since 2009, even with an additional section starting in 2013. 
You can hover over the plot to see the actual enrollment numbers. 

```{r}
p <- ggplot(class, aes(x=term_idx, y=enrollment, group=term, col=term)) + 
  geom_point() + geom_smooth(se=FALSE) + xlab("Semester") + ylab("Enrollmenet") +
  scale_color_manual(values=term.col, name = "Term")+ theme_bw() +
  scale_y_continuous(limits=c(40, 180)) + ggtitle("Class Size")

ggplotly(p, tooltip = c("y", "x")) %>% config(displayModeBar = F)
```
  

## GPA
The average class-level GPA across years is `r round(mean(class$classgpa, na.rm=TRUE), 2)`, with standard deviation `r round(sd(class$classgpa, na.rm=TRUE), 2)`. Spring and Fall semesters have different patterns of oscillating increase and decrease of class-level GPA, but there is no statistical linear trend across terms nor a difference in average GPA between Fall and Spring. 

```{r}
#lm(classgpa~as.numeric(term_idx)+ term, data=class)
#t.test(classgpa~term, data=class)

p <- ggplot(class, aes(x=term_idx, y=classgpa, group=term, col=term)) + 
  geom_point(aes(size=enrollment)) + geom_smooth(se=FALSE) + xlab("Semester") + 
  scale_color_manual(values=term.col, name = "Term") + theme_bw() +
  scale_y_continuous(limits = c(1,3)) + ylab("Class Average GPA, size of point depicts size of class") 

ggplotly(p, tooltip = c("y", "x", "group"))%>% config(displayModeBar = F)
```



## DFW rate

As expected we see a similar pattern, slightly different between Fall and Spring, for the number and % of student receiving a repeatable grade _(D/F/WU)_. 

```{r}
c_0NA <- class %>% filter(term_idx != "NA") %>% mutate(dfw_rate = round(dfw_rate,2))

a <- ggplot(class, aes(x=term_idx, y=dfw_rate, group=term, col=term)) + ylab("") + 
            geom_point(aes(size=enrollment)) + geom_smooth(se=FALSE) + xlab("Semester") + 
            scale_color_manual(values=term.col, name = "Term")+ theme_bw() + 
            scale_y_continuous(limits=c(0,1), breaks=seq(0, .80, by=.20),
                               labels = scales::percent_format()) + 
            ggtitle("DWF Rate by semester, size of point depicts size of class") 

ggplotly(a, tooltip = c("y", "size", "enrollment"))%>% config(displayModeBar = F)

```

# Student Characteristics 

## Demographics

### Gender

Campus has a seen a steady rate of slightly higher than 50% female students for the past 7 years. Female enrollment in Chem111 has been historically much lower than the campus average, but a clear uptick is being seen since AY14-15. 

```{r gender}
campus.pf <- campus_props %>% filter(var=="gender" & level=="F") %>% 
              mutate(pf = pct) %>% select(term_idx, term, pf)

pft <- chem %>% 
  filter(term_idx != "NA") %>%
  group_by(term, term_idx, section) %>% 
  summarise(pf = mean((gender=="F"), na.rm=TRUE)) %>%
  bind_rows(campus.pf) %>% mutate(percent_female = round(pf,2))

a <- ggplot(pft, aes(x=term_idx, y=percent_female, col=term, group=term)) +
  geom_point() + geom_smooth(se=FALSE) + ylab("") + 
  scale_color_manual(values=c("black", term.col), name = "Term")+ theme_bw() +
  ggtitle("Proportion of females per class") + xlab("Semester") + 
  scale_y_continuous(labels = scales::percent_format(), limits=c(.2, .7)) 


ggplotly(a, tooltip = c("y", "x", "group"))%>% config(displayModeBar = F)

rm(a, pft, campus.pf)
```


### Age
```{r age}
ave <- campus_aves %>% filter(var=="age") %>% mutate(mean=median) %>%
          select(term, term_idx, mean, p10, p90) 
            
age <- chem %>% 
  filter(term_idx != "NA") %>%
  group_by(term, term_idx, section) %>% 
  summarise(mean = mean(age, na.rm=TRUE)) %>%
  bind_rows(ave) %>% mutate(mean=round(mean,1))

p <- ggplot(age, aes(x=term_idx, y=mean, col=term, group=term)) +
      geom_point(size=3) + geom_smooth(se=FALSE) + ylab("Age") + 
      scale_color_manual(values=c("grey20", term.col), name = "Term")+ theme_bw() +
      ggtitle("Average student age per section") + xlab("Semester") + 
      scale_y_continuous(limits=c(17, 24))
  
ggplotly(p, tooltip = c("y", "x", "group")) %>% config(displayModeBar = F)
rm(ave, age, p)
```
  
The black horizontal line represents the **median** age of all students across campus -- at the time of enrollment. From 2009 to about 2013 there was an increase in the average age of students at enrollment campus wide, which has seemed to have leveled off. 

Students taking Chemistry tend to be slightly younger than the average campus age, and those taking Chemistry in the fall are on average younger than those taking chemistry in the Spring. Recall this is age at admission, not the age when the student is taking Chem 111. 


### Where the stuent comes from

The patterns of trend for the shifting demographics for where the students come from and enroll in Chemistry is similar to the overall campus trends. We see a decline in the proportion of students coming from Butte County (gold line), our general service area (red-orange line), and from the Rest of California (purple line) in general. There is steep increase in the enrollment from students in the Los Angeles area since around Fall 2011 (green line). By the F16 23% of the campus population is from the LA area, whereas these students make up 31% of Chemistry 111 classes in F16. 

```{r resid}
rt <- chem %>%
  filter(residareaname != "NA", term_idx != "NA") %>%
  group_by(term_idx, residareaname) %>%
  summarise(n=n()) %>%
  mutate(pct=round(n/sum(n),2), term="Chem 111")

campus.area <- campus_props %>% filter(var=="residareaname") %>% 
                mutate(residareaname=level) %>% select(term_idx, term, residareaname, pct)

art <- bind_rows(rt, campus.area)

ggplot(art, aes(x=term_idx, y=pct, col=residareaname, group=residareaname)) + geom_point() +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0,.5)) + xlab("Semester") +
  ylab("% Areas") + geom_smooth(se=FALSE) + theme_bw() + theme(legend.position="bottom") + 
  scale_color_discrete(name = "Area of Residence") + facet_wrap(~term)
```

### Race / Ethnicity
For completeness sake, we first examine the trends for all recorded race/ethnicity (IPEDS) classifications. 
The most notable trends is the decrease in White students from being the majority at 59.6% in Fall 09 down to 44.1% by Fall 2016, with an equivalent increase in the proportion of students who identify as Hispanic/Latino increasing from 16.4% in Fall 2009 to 32.5% by Fall 2016. 


In 2009, the proportion of White and Hispanic/Latino students taking Chemistry was similar to the campus averages (53.2% White and 17.7% Hispanic/Latino). However the proportion of Hispanic/Latino students rose to match that of White students taking Chemistry 111 in Fall 2015: 40.5% white and 40.1% Hispanic/Latino. This is a more equal balance compared to the campus average in Fall 2016 which is 44.1% White and 32.5% Hispanic/Latino. 

```{r race}
rct <- chem %>%
  filter(race_code != "NA", term_idx != "NA") %>%
  group_by(term_idx, race_code) %>%
  summarise(n=n()) %>%
  mutate(pct=n/sum(n), term="Chemistry 111")

campus.area <- campus_props %>% filter(var=="race_code") %>% 
                mutate(race_code=level) %>% select(term_idx, term, race_code, pct)

art <- bind_rows(rct, campus.area)

ggplot(art, aes(x=term_idx, y=pct, col=race_code, group=race_code)) + geom_point() +
  xlab("Semester") +  ylab(" ") +
  geom_smooth(se=FALSE) + theme_bw()  + theme(legend.position="bottom") +
  scale_y_continuous(limits=c(0, .75), labels = scales::percent_format()) +
  scale_color_discrete(name = "Race / Ethnicity") + facet_wrap(~term)
```
  
Since eight race/ethnic categories is too many to compare, especially when six of those categories make up less than 80% of the population, two additional groupings of categories are commonly used: One to identify "underrepresented minority" students, and the other to identify "students of color". The historical trends for both of these are looked at next, but essentially show the same information as the plot just discussed. 


### URM
The definition used in this report follows the campus definition for an underrepresented minority student and is as follows:

* Non-URM: Asian, NHOPI, White
* URM: Black, HL
* Unknown: Mult, Unk

Since 2009, Chemistry 111 has had a higher proportion of URM students enrolled compared to the campus average. This gap has widened since around Fall 2012 or 2013, the proportion of URM taking in Chem 111 is increasing faster than the campus average. 


```{r urm}
campus.purm <- campus_props %>% filter(var=="urm_status") %>% mutate(urm_status=level) %>% 
                  select(term_idx, urm_status, pct, term)

urmt <- chem %>%
  filter(urm_status != "NA", term_idx != "NA") %>%
  group_by(term_idx, urm_status) %>%
  summarise(n=n()) %>%
  mutate(pct=n/sum(n), term="Chemistry 111") %>%
  bind_rows(campus.purm)


ggplot(urmt, aes(x=term_idx, y=pct, col=interaction(urm_status, term), 
                 group=interaction(urm_status, term), shape=term)) + 
  geom_point() + geom_smooth(se=FALSE) + theme_bw() + 
  xlab("Semester") +  ylab("% URM status") +
  scale_y_continuous(limits=c(0, .75), labels = scales::percent_format()) +
  scale_shape_discrete(name = " ") +
  scale_colour_manual(name = "URM Status"
                      , values=brewer.pal(6, "Paired")[c(1,3,5,2,4,6)], 
                      labels=c("Non-URM: Campus Average",
                                 "Unknown: Campus Average", 
                                 "URM: Campus Average", 
                                 "Non-URM: Chemistry 111", 
                                 "Unknown: Chemistry 111", 
                                 "URM: Chemistry 111")) 

rm(urmt, campus.purm)
```
  
### SOC

The definition used in this report for "Student of Color" is as follows:

* SOC: American Indian/Alaskan Native, Asian, Black, Hispanic/Latino, 
* Non-SOC: White
* Multiple/Mixed: Multiple, Unknown

A very similar trend is seen using the classification that lumps all students non of white or of mixed or unknown ethnicity together. Using this metric, the proportion of students of color surpassed the proportion of white students in Chemistry 111 by Spring 2014. 
By Fall 2016 students taking Chemistry 111 are 47.7% students of color and 38.6% White

```{r soc}

campus.psoc <- campus_props %>% filter(var=="soc") %>% mutate(soc=level) %>% 
                  select(term_idx, soc, pct, term)

soct <- chem %>%
  filter(soc != "NA", term_idx != "NA") %>%
  group_by(term_idx, soc) %>%
  summarise(n=n()) %>%
  mutate(pct=n/sum(n), term="Chemistry 111") %>%
  bind_rows(campus.psoc)


ggplot(soct, aes(x=term_idx, y=pct, col=interaction(soc, term), 
                 group=interaction(soc, term), shape=term)) + 
  geom_point() + geom_smooth(se=FALSE) + theme_bw() + 
  xlab("Semester") +  ylab(" ") +
  scale_y_continuous(limits=c(0, .75), labels = scales::percent_format()) +
  scale_shape_discrete(name = " ") +
  scale_colour_manual(name = "Student of Color"
                      , values=brewer.pal(6, "Paired")[c(1,3,5,2,4,6)]
                      , labels=c("Mixed: Campus Average", 
                                 "White: Campus Average", 
                                 "SOC: Campus Average", 
                                 "Mixed: Chemistry 111", 
                                 "White: Chemistry 111", 
                                 "SOC: Chemistry 111")) 

rm(soct, campus.psoc)

```

### First Generation
There has been a steady and statistically significant increase in the proportion of first generation students attending Chico State, and enrolling in Chemistry 111 over the past seven years. In this case, the Fall Chemistry student body is no different than the average Chico State student, but since about 2013 the proportion of first generation students in the Spring cohort has been increasing.  

```{r firstgen}
#summary(lm(pfg~as.numeric(term_idx)+ term, data=pfgt))

campus.pfg <- campus_props %>% filter(var=="firstgeneration" & level=="Y") %>% 
              mutate(pfg = pct) %>% select(term_idx, term, pfg)

pfgt <- chem %>% 
  filter(term_idx != "NA") %>%
  group_by(term, term_idx, section) %>% 
  summarise(pfg = mean((firstgeneration=="Y"), na.rm=TRUE)) %>%
  bind_rows(campus.pfg) %>% mutate(pct_firstgen = round(pfg,2))

a <- ggplot(pfgt, aes(x=term_idx, y=pct_firstgen, col=term, group=term)) +
  geom_point() + geom_smooth(se=FALSE) + ylab(" ") + 
  scale_color_manual(values=c("black", term.col), name = "Term")+ theme_bw() +
  xlab("Semester") + 
  scale_y_continuous(labels = scales::percent_format(), limits=c(.2, .7)) 

ggplotly(a, tooltip = c("y", "x", "group"))%>% config(displayModeBar = F)


rm(a, pfgt, campus.pfg)
```
  

### Citizenship
The proportion of non-citizens enrolled campus-wide levels off at about 5%, but there has been an increase in non-citizens enrolling in Chem 111 since around 2011. The three sections of Chem 111 in Fall 16 had 5%, 8% and 11% non-citizens enrolled in their classes.

```{r citizen}
campus.pnc <- campus_props %>% filter(var=="citizen" & level=="N") %>% 
              mutate(pnc = pct) %>% select(term_idx, term, pnc)
ct <- chem %>%
  filter(term_idx != "NA") %>%
  group_by(term, term_idx, section) %>% 
  summarise(pnc = mean((citizen=="N"), na.rm=TRUE)) %>%
  bind_rows(campus.pnc) %>% mutate(percent_noncitizen = round(pnc,2))

b <- ggplot(ct, aes(x=term_idx, y=percent_noncitizen, col=term, group=term)) +
  geom_point() + geom_smooth(se=FALSE) + ylab(" ") + 
  scale_color_manual(values=c("black", term.col), name = "Term")+ theme_bw() +
  xlab("Semester") +
  scale_y_continuous(limits=c(0, .25), labels = scales::percent_format())

ggplotly(b, tooltip = c("y", "x", "group")) %>% config(displayModeBar = F)

rm(b, campus.pnc, ct)
```
  
  
### Disabled Student Services

Students who are receiving services from DSS has varied over the past 7 years, but generally stayed below 5%, which is what the campus average has been at since around Spring 2013. 

```{r dss}
campus.pd <- campus_props %>% filter(var=="dss" & level=="Y") %>% 
              mutate(pd = pct) %>% select(term_idx, term, pd)
dt <- chem %>%
  filter(term_idx != "NA") %>%
  group_by(term, term_idx, section) %>% 
  summarise(pd = mean((DSS=="DSS"), na.rm=TRUE)) %>%
  bind_rows(campus.pd) %>% mutate(percent_dss = round(pd,2))
cc <- ggplot(dt, aes(x=term_idx, y=percent_dss, col=term, group=term)) +
  geom_point() + geom_smooth(se=FALSE) + ylab(" ") + 
  scale_color_manual(values=c("black", term.col), name = "Term")+ theme_bw() +
  ggtitle("Proportion of DSS students per section") + xlab("Semester") +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, .25))

ggplotly(cc, tooltip = c("y", "x", "group"))%>% config(displayModeBar = F)

rm(cc, dt, campus.pd)
```
  
  
### EOP Status
The proportion of students who are participating in the EOP program while enrolled in Chem 111 have historically been higher than the campus level average. There is indication that while the campus average remains constant at about 7%, both Fall and Spring terms for Chemistry have declined to 3-4% by 2016. Between 2011 and 2014 the average proportion of students in EOP was higher in Spring than it was in the Fall, nearly double in some cases (6% vs 12%).

**Note** These higher numbers of EOP students in the spring may be students who did not pass Chem 111 in the Fall and so are repeating the class in the Spring. The analysis conducted in this report does not examine this effect directly. 

```{r eop}
campus.peop <- campus_props %>% filter(var=="eopstat" & level=="Y") %>% 
              mutate(peop = pct) %>% select(term_idx, term, peop)

eopt <- chem %>%
  filter(term_idx != "NA") %>%
  group_by(term, term_idx, section) %>% 
  summarise(peop = mean(eop, na.rm=TRUE)) %>%
  bind_rows(campus.peop) %>% mutate(percent_eop = round(peop,2))

d <- ggplot(eopt, aes(x=term_idx, y=percent_eop, col=term, group=term)) +
  geom_point() + geom_smooth(se=FALSE) + ylab(" ") + 
  scale_color_manual(values=c("black", term.col), name = "Term")+ theme_bw() +
  ggtitle("Proportion of EOP students per section") + xlab("Semester") +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, .25))

 campus.pd <- campus_props %>% filter(var=="dss" & level=="Y") %>% 
              mutate(pd = pct) %>% select(term_idx, term, pd)

ggplotly(d, tooltip = c("y", "x", "group"))%>% config(displayModeBar = F)

rm(d, eopt, campus.pd)
```
  


### REACH

Data on the reach program was was either not contributing data to the SIS system or not in place prior to Fall 2013.
Campus level averages for the Reach program were also not available. Similar to EOP, the proportion of students in the REACH program are higher in Spring compared to Fall. However it is my opinion that this data is not reliable due to some discrepencies found while working with Erik. 

```{r reach}
reacht <- chem %>%
  filter(term_idx != "NA") %>%
  group_by(term, term_idx, section) %>% 
  summarise(prea = mean(reach, na.rm=TRUE))

d <- ggplot(reacht, aes(x=term_idx, y=prea, col=term, group=term)) +
  geom_point() + geom_smooth(se=FALSE) + ylab(" ") + 
  scale_color_manual(values=term.col, name = "Term") + theme_bw() +
  ggtitle("Proportion of Reach students per section") + xlab("Semester") +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, .25))

ggplotly(d, tooltip = c("y", "x", "group"))%>% config(displayModeBar = F)

rm(d, reacht)
```
  
  
## Pre-enrollment Information  
  
### Admission Index
Students enrolling in Chemistry 111 tend to have a higher admissions index (high school eligibility index (HSEI)) than the average student on campus. This average is created based on all students on campus during that semester, but the HSEI measure is only created upon admission. 

There was a lot of variation in the average index prior to around 2013, the campus level HSEI in Spring 10 was 3.44, up to 3.66 in Fall 10, back down to 3.26 in Spring 11, and up again to 3.60 in the Fall. 

There may be a slight upward trend for both Chemistry student and the campus average, after 2014 we see class level averages for Chemsitry courses that are over 3.8. 


```{r hsei}
ave <- campus_aves %>% filter(var=="admission_index") %>% 
          select(term, term_idx, mean) 
            
ai <- chem %>% 
  filter(term_idx != "NA") %>%
  group_by(term, term_idx, section) %>% 
  summarise(mean = mean(admission_index, na.rm=TRUE)) %>%
  bind_rows(ave) %>% mutate(mean=round(mean,2))

a <- ggplot(ai, aes(x=term_idx, y=mean, col=term, group=term)) +
      geom_point(size=3) + geom_smooth(se=FALSE) + ylab("Admissions Index") + 
      scale_color_manual(values=c("grey20", term.col), name = "Term")+ theme_bw() +
      xlab("Semester") + 
      scale_y_continuous(limits=c(3, 4))
  
ggplotly(a, tooltip = c("y", "x", "group")) %>% config(displayModeBar = F)

rm(ai, ave, a)
```

```{r, eval=FALSE}
hsei_out <- biv.var(df=chem, var='admission_index',  row.var='section', col.var='term_idx')
hsei_out1 <- hsei_out[1:3,1:8]
hsei_out2 <- hsei_out[c(1,2,4),c(1,9:15)]

kable(hsei_out1,  align='c', row.names=FALSE, caption='Mean(sd) HSEI per section, per term (09-12)')
kable(hsei_out2,  align='c', row.names=FALSE, caption='Mean(sd) HSEI per section, per term (13-16)')
```



  
## Tests and Benchmarks

### GE-Breadth Course Completion Status

```{r ge}
matht <- chem %>%
  filter(term_idx != "NA") %>%
  group_by(term_idx, section) %>% 
  summarise(prop = mean(gemathdone, na.rm=TRUE)) %>%
  mutate(GE = "Math")
engt <- chem %>%
  filter(term_idx != "NA") %>%
  group_by(term_idx, section) %>% 
  summarise(prop = mean(geenlgdone, na.rm=TRUE)) %>%
  mutate(GE = "English")  
critt <- chem %>%
  filter(term_idx != "NA") %>%
  group_by(term_idx, section) %>% 
  summarise(prop = mean(gecritdone, na.rm=TRUE)) %>%
  mutate(GE = "Critical Thinking")
get <- rbind(matht, engt, critt)

ggplot(get, aes(x=term_idx, y=prop, col=GE, group=GE)) + geom_point() + xlab("Semester") + 
  ggtitle("Proportion of students who have completed these GE's\nprior to enrolling at a CSU campus per semester") +
  ylab("% Completed GE's") + geom_smooth(se=FALSE) + theme_bw() +
  scale_y_continuous(limits=c(0, .5), labels = scales::percent_format())

```
  

### ELM
```{r elm}
campus.pelm <- campus_props %>% filter(var=="elm" & level=="1") %>% 
              mutate(pelm = pct) %>% select(term_idx, term, pelm)

elmt <- chem %>%
  filter(term_idx != "NA") %>%
  group_by(term, term_idx, section) %>% 
  summarise(pelm = mean(elm, na.rm=TRUE)) %>%
  bind_rows(campus.pelm) %>% mutate(percent_elm = round(pelm,2))

e <- ggplot(elmt, aes(x=term_idx, y=percent_elm, col=term, group=term)) +
  geom_point() + geom_smooth(se=FALSE) + ylab("") + 
  scale_color_manual(values=c("black", term.col), name = "Term") + theme_bw() +
  ggtitle("Proportion of students per section who have completed\nthe ELM requirement prior to enrollment in CHEM") +
  xlab("Semester") +
  scale_y_continuous(labels = scales::percent_format(), limits=c(.75,1))

ggplotly(e, tooltip = c("y", "x", "group"))%>% config(displayModeBar = F)

rm(e, elmt, campus.pelm)
```
  

### EPT

```{r ept}
campus.pept <- campus_props %>% filter(var=="ept" & level=="1") %>% 
              mutate(pept = pct) %>% select(term_idx, term, pept)
eptt <- chem %>%
  filter(term_idx != "NA") %>%
  group_by(term, term_idx, section) %>% 
  summarise(pept = mean(ept, na.rm=TRUE)) %>%
  bind_rows(campus.pept) %>% mutate(percent_ept = round(pept,2))

f <- ggplot(eptt, aes(x=term_idx, y=percent_ept, col=term, group=term)) +
  geom_point() + geom_smooth(se=FALSE) + ylab("") + 
  scale_color_manual(values=c("black", term.col), name = "Term") + theme_bw() +
  ggtitle("Proportion of students per section who have completed\nthe EPT requirement prior to enrollment in CHEM") +
  xlab("Semester") +
  scale_y_continuous(labels = scales::percent_format(), limits = c(.8, 1))

ggplotly(f, tooltip = c("y", "x", "group"))%>% config(displayModeBar = F)

rm(f, eptt, campus.pept)
```
  

  
### College prep

    

    
## Academic Standing

### College
```{r college}
rt <- chem %>%
  group_by(term_idx, college_name) %>%
  summarise(n=n()) %>%
  mutate(pct=n/sum(n))
ggplot(rt, aes(x=term_idx, y=pct, col=college_name, group=college_name)) + geom_point() +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0,.5)) + xlab("Semester") + ylab(" ") +
  geom_smooth(se=FALSE) + theme_bw() + scale_color_discrete(name = "College Name")
```
  
  
### Student Level
```{r studlvl}
slt <- chem %>%
  filter(!is.na(stulevel)) %>%
  group_by(term_idx, stulevel) %>%
  summarise(n=n()) %>%
  mutate(percent=round(n/sum(n),2))

a <- ggplot(slt, aes(x=term_idx, y=percent, col=stulevel, group=stulevel)) + geom_point() +
      scale_y_continuous(labels = scales::percent_format(), limits = c(0,.75)) + xlab("Semester") +
      ggtitle("Proportion of students at each academic status per semester") + ylab("") +
      geom_smooth(se=FALSE) + theme_bw() + 
      scale_colour_discrete(name = "Student Level",
                            labels = c("Freshman", "Sophomore", "Junior", "Senior"))

ggplotly(a, tooltip = c("y", "x", "group"))%>% config(displayModeBar = F)

rm(a, slt)
```

  

### Units taken this semester

```{r units}
utt <- chem %>% 
  filter(term_idx != "NA") %>%
  group_by(term, term_idx, section) %>% 
  summarise(mean = mean(term_units, na.rm=TRUE))

ggplot(utt, aes(x=term_idx, y=mean, col=term, group=term)) +
  geom_point() + geom_smooth(se=FALSE) + ylab("Average units") + 
  scale_color_manual(values=term.col, name = "Term")+ theme_bw() +

  xlab("Semester") + scale_y_continuous()
```
  
  



