---
title: 'Historical Trends in Students taking Chem 111'
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

This section analyses the trends in characteristics of the student group who is taking Chemistry 111, and compares them to the campus level averages for all freshman. 

We consider data from Fall 2009 - Fall 2016, for all sections of Chemistry 111. Students withdrawing or receving an incomplete (`W/I`) are exluded from this analysis. A repeatable grade is defined as a student reciving a `D, D+`, or `F` grades, or an unexcused withdrawl `WU`. 


```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr);library(knitr); library(RColorBrewer)
library(ggplot2);library(gridExtra); library(grid)
library(scales); library(foreach); library(plotly)

'%ni%' <- Negate('%in%')

opts_chunk$set(warning=FALSE, message=FALSE, echo=FALSE, row.names = FALSE, cache=FALSE,
               fig.align = 'center',  fig.width=8, fig.height=5)
options(knitr.kable.NA = '') 

source("C:/Dropbox/Professional/Grants_Current/SS Chemistry Data Mining/code/helper functions.R")
rawtrends <-read.delim("C:/Dropbox/Professional/Grants_Current/SS Chemistry Data Mining/data/chem_data_20170227.txt", sep = "\t")
campus_aves <- read.delim("C:/Dropbox/Professional/Grants_Current/SS Chemistry Data Mining/data/campus_aves.txt", sep="\t")
campus_props <- read.delim("C:/Dropbox/Professional/Grants_Current/SS Chemistry Data Mining/data/campus_props.txt", sep="\t")


chem <- rawtrends %>% filter(grade %ni% c('W ', 'I ') & erss_stulevel < 5) %>% droplevels() %>%
            mutate(term_idx = factor(term_idx, 
                      levels = c("F09", "S10", "F10", "S11", "F11", "S12", "F12", "S13", 
                                "F13", "S14", "F14", "S15", "F15", "S16", "F16")),
                    stulevel = factor(erss_stulevel, labels = c("Freshman", "Sophomore", "Junior", "Senior")), 
                    DSS = ifelse(dss.1==1 | dss == 'Y', 'DSS', 'not in DSS'), 
                    repeat_grade = ifelse(grade %in% c('D ', 'D+', 'F ', 'WU'), 1, 0)) %>%
            select(-dss.1, -dss, -erss_stulevel, erss_enrstat, -dfw_rate)
                

# redo class level measures
class <- chem %>% group_by(year, term, term_idx, section) %>%
          summarise(classgpa = mean(gpa), dfw = sum(repeat_grade), 
                    enrollment = n(), dfw_rate = dfw/enrollment)
         
yearlab <- c(2009, rep(2010:2016, each=2))
term.col <- brewer.pal(9, "Set1")[c(7, 3)]
```


# Course Characteristics

## Enrollment
```{r}
ggplot(chem, aes(x=term_idx, y=..count.., fill=term)) + geom_bar() +
  geom_text(stat="count", aes(y = ..count.., vjust=-.3, label=..count..)) +
  xlab("Year") + ylab("Enrollment") + theme_bw() +  
  scale_x_discrete(labels=yearlab) + scale_fill_manual(values=term.col, name = "Term")
```
  
Spring term demand has steadily increased since 2009. 
 
## Class size

```{r}
p <- ggplot(class, aes(x=term_idx, y=enrollment, group=term, col=term)) + 
  geom_point() + geom_smooth(se=FALSE) + xlab("Semester") + ylab("Enrollmenet") +
  scale_color_manual(values=term.col, name = "Term")+ theme_bw() +
  scale_y_continuous(limits=c(40, 180)) + ggtitle("Class Size")

ggplotly(p, tooltip = c("y", "x"))
```
  
Fall class sizes have remained relativley constant between 140-160 since 2009, even with an additional section starting in 2013. 


## GPA
The average class-level gpa across years is `r round(mean(class$classgpa, na.rm=TRUE), 2)`, with standard deviation `r round(sd(class$classgpa, na.rm=TRUE), 2)`. 

```{r}
p <- ggplot(class, aes(x=term_idx, y=classgpa, group=term, col=term)) + 
  geom_point() + geom_smooth(se=FALSE) + xlab("Semester") + 
  scale_color_manual(values=term.col, name = "Term") + theme_bw() +
  scale_y_continuous(limits = c(1,3)) + ylab("Class Average GPA") 

ggplotly(p, tooltip = c("y", "x", "group"))
```

Spring and Fall semesters have different patterns of oscilating increase and decrease of class-level GPA. 


## DFW rate

As expected we see a similar pattern, slightly different between Fall and Spring, for the number and % of student receiving a repeatable grade _(D/F/WU)_. 

```{r}
c_0NA <- class %>% filter(term_idx != "NA") %>% mutate(dfw_rate = round(dfw_rate,2))

a <- ggplot(class, aes(x=term_idx, y=dfw_rate, group=term, col=term)) + ylab("") + 
            geom_point(aes(size=enrollment)) + geom_smooth(se=FALSE) + xlab("Semester") + 
            scale_color_manual(values=term.col, name = "Term")+ theme_bw() + 
            scale_y_continuous(limits=c(0,1), breaks=seq(0, .80, by=.20),
                               labels = scales::percent_format()) + 
            ggtitle("DWF Rate by semester, size of point depicts size of class") 

ggplotly(a, tooltip = c("y", "size", "enrollment"))

```

# Student Characteristics 

## Demographics

### Gender
  
```{r}
campus.pf <- campus_props %>% filter(var=="gender" & level=="F") %>% 
              mutate(pf = pct) %>% select(term_idx, term, pf)

pft <- chem %>% 
  filter(term_idx != "NA") %>%
  group_by(term, term_idx, section) %>% 
  summarise(pf = mean((gender=="F"), na.rm=TRUE)) %>%
  bind_rows(campus.pf) %>% mutate(percent_female = round(pf,2))

a <- ggplot(pft, aes(x=term_idx, y=percent_female, col=term, group=term)) +
  geom_point() + geom_smooth(se=FALSE) + ylab("") + 
  scale_color_manual(values=c("black", term.col), name = "Term")+ theme_bw() +
  ggtitle("Proportion of females per class") + xlab("Semester") + 
  scale_y_continuous(labels = scales::percent_format(), limits=c(.2, .7)) 


ggplotly(a, tooltip = c("y", "x", "group"))

rm(a, pft, campus.pf)
```

Campus has a seen a steady rate of slightly higher than 50% female students for the past 7 years. 

 
### Age
```{r}
ave <- campus_aves %>% filter(var=="age") %>% mutate(mean=median) %>%
          select(term, term_idx, mean, p10, p90) 
            
age <- chem %>% 
  filter(term_idx != "NA") %>%
  group_by(term, term_idx, section) %>% 
  summarise(mean = mean(age, na.rm=TRUE)) %>%
  bind_rows(ave) %>% mutate(mean=round(mean,1))

p <- ggplot(age, aes(x=term_idx, y=mean, col=term, group=term)) +
      geom_point() + geom_smooth(se=FALSE) + geom_pointrange(aes(ymin=p10, ymax=p90))+ 
      ylab("Age") + 
      scale_color_manual(values=c("black", term.col), name = "Term")+ theme_bw() +
      ggtitle("Average student age per section") + xlab("Semester") + 
      scale_y_continuous(limits=c(17, 30))
  
ggplotly(p, tooltip = c("y", "x", "group"))
rm(ave, age, p)
```
  
The black horizonal line represents the **median** age of freshman across campus, with vertical lines spanning the youngest 10% to the oldest 90% of ages for freshman that term. Students taking Chemistry are on average older than most freshman -- this is expected because there are a proportion of second year students taking Chemistry that will increase the average age of the class. 

Students taking Chemistry in the fall are on average younger than those taking chemistry in the Spring. The obvious exception is this past Spring 2016 seems to be an unusually young cohort for spring. Recall this is age at admission, not the age when the student is taking Chem 111. 



### Area Where a Student is From
```{r}
rt <- chem %>%
  filter(residareaname != "NA", term_idx != "NA") %>%
  group_by(term_idx, residareaname) %>%
  summarise(n=n()) %>%
  mutate(pct=round(n/sum(n),2))

a <- ggplot(rt, aes(x=term_idx, y=pct, col=residareaname, group=residareaname)) + geom_point() +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0,.5)) + xlab("Semester") +
  ggtitle("Proportion of areas where students are from per semester") + ylab("% Areas") +
  geom_smooth(se=FALSE) + theme_bw() + scale_color_discrete(name = "Area of Residence")

ggplotly(a, tooltip = c("y", "group"))

```

### Citizenship

```{r}
campus.pnc <- campus_props %>% filter(var=="citizen" & level=="N") %>% 
              mutate(pnc = pct) %>% select(term_idx, term, pnc)
ct <- chem %>%
  filter(term_idx != "NA") %>%
  group_by(term, term_idx, section) %>% 
  summarise(pnc = mean((citizen=="N"), na.rm=TRUE)) %>%
  bind_rows(campus.pnc) %>% mutate(percent_noncitizen = round(pnc,2))

b <- ggplot(ct, aes(x=term_idx, y=percent_noncitizen, col=term, group=term)) +
  geom_point() + geom_smooth(se=FALSE) + ylab("") + 
  scale_color_manual(values=c("black", term.col), name = "Term")+ theme_bw() +
  ggtitle("Proportion of students that are not citizens per section") + xlab("Semester") +
  scale_y_continuous(limits=c(0, .25), labels = scales::percent_format())
ggplotly(b, tooltip = c("y", "x", "group"))

rm(b, campus.pnc, ct)
```
  
  
### Students who are recieving services from DSS
```{r}
campus.pd <- campus_props %>% filter(var=="dss" & level=="Y") %>% 
              mutate(pd = pct) %>% select(term_idx, term, pd)
dt <- chem %>%
  filter(term_idx != "NA") %>%
  group_by(term, term_idx, section) %>% 
  summarise(pd = mean((DSS=="DSS"), na.rm=TRUE)) %>%
  bind_rows(campus.pd) %>% mutate(percent_dss = round(pd,2))
cc <- ggplot(dt, aes(x=term_idx, y=percent_dss, col=term, group=term)) +
  geom_point() + geom_smooth(se=FALSE) + ylab("") + 
  scale_color_manual(values=c("black", term.col), name = "Term")+ theme_bw() +
  ggtitle("Proportion of DSS students per section") + xlab("Semester") +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, .1))

ggplotly(cc, tooltip = c("y", "x", "group"))

rm(cc, dt, campus.pd)
```
  
  
### EOP Status
```{r}
campus.peop <- campus_props %>% filter(var=="eopstat" & level=="Y") %>% 
              mutate(peop = pct) %>% select(term_idx, term, peop)

eopt <- chem %>%
  filter(term_idx != "NA") %>%
  group_by(term, term_idx, section) %>% 
  summarise(peop = mean(eop, na.rm=TRUE)) %>%
  bind_rows(campus.peop) %>% mutate(percent_eop = round(peop,2))

d <- ggplot(eopt, aes(x=term_idx, y=percent_eop, col=term, group=term)) +
  geom_point() + geom_smooth(se=FALSE) + ylab("") + 
  scale_color_manual(values=c("black", term.col), name = "Term")+ theme_bw() +
  ggtitle("Proportion of EOP students per section") + xlab("Semester") +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, .25))

 campus.pd <- campus_props %>% filter(var=="dss" & level=="Y") %>% 
              mutate(pd = pct) %>% select(term_idx, term, pd)

ggplotly(d, tooltip = c("y", "x", "group"))

rm(d, eopt, campus.pd)
```
  


### REACH
    
```{r}
reacht <- chem %>%
  filter(term_idx != "NA") %>%
  group_by(term, term_idx, section) %>% 
  summarise(prea = mean(reach, na.rm=TRUE))
ggplot(reacht, aes(x=term_idx, y=prea, col=term, group=term)) +
  geom_point() + geom_smooth(se=FALSE) + ylab("% Reach") + 
  scale_color_manual(values=term.col, name = "Term") + theme_bw() +
  ggtitle("Proportion of Reach students per section") + xlab("Semester") +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, .25))
```
  
### Any formal student group (fsg)
  
```{r}
fsgt <- chem %>%
  filter(term_idx != "NA") %>%
  group_by(term, term_idx, section) %>% 
  summarise(pfsg = mean(fsg, na.rm=TRUE))
ggplot(fsgt, aes(x=term_idx, y=pfsg, col=term, group=term)) +
  geom_point() + geom_smooth(se=FALSE) + ylab("% FSG") + 
  scale_color_manual(values=term.col, name = "Term") + theme_bw() +
  ggtitle("Proportion of students in any FSG per section") + xlab("Semester") +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, .5))
```
  
### <a id="race_trends"></a> Race Code 
```{r}
rct <- chem %>%
  filter(race_code != "NA", term_idx != "NA") %>%
  group_by(term_idx, race_code) %>%
  summarise(n=n()) %>%
  mutate(prc=n/sum(n))
ggplot(rct, aes(x=term_idx, y=prc, col=race_code, group=race_code)) + geom_point() +
  xlab("Semester") + ggtitle("Proportion of students' race per semester") + ylab("% RACE") +
  geom_smooth(se=FALSE) + theme_bw() + scale_y_continuous(limits=c(0, 1), labels = scales::percent_format()) +
  scale_color_discrete(name = "Race")
```
  


### URM

  
```{r}
urmt <- chem %>%
  filter(urm_status != "NA", term_idx != "NA") %>%
  group_by(term_idx, urm_status) %>%
  summarise(n=n()) %>%
  mutate(purm=n/sum(n))
ggplot(urmt, aes(x=term_idx, y=purm, col=urm_status, group=urm_status)) + geom_point() +
  xlab("Semester") + ggtitle("Proportion of students' URM status per semester") + ylab("% URM status") +
  geom_smooth(se=FALSE) + theme_bw() + scale_y_continuous(limits=c(0, 1), labels = scales::percent_format()) +
  scale_color_discrete(name = "URM Status")

```
  
### SOC

```{r}
soct <- chem %>%
  filter(soc != "NA", term_idx != "NA") %>%
  group_by(term_idx, soc) %>%
  summarise(n=n()) %>%
  mutate(psoc=n/sum(n))
ggplot(soct, aes(x=term_idx, y=psoc, col=soc, group=soc)) + geom_point() +
  xlab("Semester") + ggtitle("Proportion of students' SOC status per semester") + ylab("% SOC status") +
  geom_smooth(se=FALSE) + theme_bw() + scale_y_continuous(limits=c(0, 1), labels = scales::percent_format()) +
  scale_color_discrete(name = "SOC Status")
```

## Pre-enrollment Information  
  
### Admission Index
```{r}
ai <- chem %>% 
  filter(term_idx != "NA") %>%
  group_by(term, term_idx, section) %>% 
  summarise(mean = mean(admission_index, na.rm=TRUE)) %>% ungroup()

aai <- mean(chem$admission_index, na.rm=TRUE)

ggplot(ai, aes(x=term_idx, y=mean, col=term, group=term)) +
  geom_point() + geom_smooth(se=FALSE) + ylab("Average admission index") + 
  scale_color_manual(values=term.col, name = "Term") + theme_bw() +
  ggtitle("Average student admission index per section") + xlab("Semester") + 
  geom_hline(yintercept=aai) + scale_y_continuous(limits=c(3.5,4))
```

```{r}
hsei_out <- biv.var(df=chem, var='admission_index',  row.var='section', col.var='term_idx')
hsei_out1 <- hsei_out[1:3,1:8]
hsei_out2 <- hsei_out[c(1,2,4),c(1,9:15)]

kable(hsei_out1,  align='c', row.names=FALSE, caption='Mean(sd) HSEI per section, per term (09-12)')
kable(hsei_out2,  align='c', row.names=FALSE, caption='Mean(sd) HSEI per section, per term (13-16)')
```



  
## Tests and Benchmarks

### GE-Breadth Course Completion Status

```{r}
matht <- chem %>%
  filter(term_idx != "NA") %>%
  group_by(term_idx, section) %>% 
  summarise(prop = mean(gemathdone, na.rm=TRUE)) %>%
  mutate(GE = "Math")
engt <- chem %>%
  filter(term_idx != "NA") %>%
  group_by(term_idx, section) %>% 
  summarise(prop = mean(geenlgdone, na.rm=TRUE)) %>%
  mutate(GE = "English")  
critt <- chem %>%
  filter(term_idx != "NA") %>%
  group_by(term_idx, section) %>% 
  summarise(prop = mean(gecritdone, na.rm=TRUE)) %>%
  mutate(GE = "Critical Thinking")
get <- rbind(matht, engt, critt)

ggplot(get, aes(x=term_idx, y=prop, col=GE, group=GE)) + geom_point() + xlab("Semester") + 
  ggtitle("Proportion of students who have completed these GE's\nprior to enrolling at a CSU campus per semester") +
  ylab("% Completed GE's") + geom_smooth(se=FALSE) + theme_bw() +
  scale_y_continuous(limits=c(0, .5), labels = scales::percent_format())

```
  

### ELM
```{r}
campus.pelm <- campus_props %>% filter(var=="elm" & level=="1") %>% 
              mutate(pelm = pct) %>% select(term_idx, term, pelm)

elmt <- chem %>%
  filter(term_idx != "NA") %>%
  group_by(term, term_idx, section) %>% 
  summarise(pelm = mean(elm, na.rm=TRUE)) %>%
  bind_rows(campus.pelm) %>% mutate(percent_elm = round(pelm,2))

e <- ggplot(elmt, aes(x=term_idx, y=percent_elm, col=term, group=term)) +
  geom_point() + geom_smooth(se=FALSE) + ylab("") + 
  scale_color_manual(values=c("black", term.col), name = "Term") + theme_bw() +
  ggtitle("Proportion of students per section who have completed\nthe ELM requirement prior to enrollment in CHEM") +
  xlab("Semester") +
  scale_y_continuous(labels = scales::percent_format(), limits=c(.75,1))

ggplotly(e, tooltip = c("y", "x", "group"))

rm(e, elmt, campus.pelm)
```
  

### EPT

```{r}
campus.pept <- campus_props %>% filter(var=="ept" & level=="1") %>% 
              mutate(pept = pct) %>% select(term_idx, term, pept)
eptt <- chem %>%
  filter(term_idx != "NA") %>%
  group_by(term, term_idx, section) %>% 
  summarise(pept = mean(ept, na.rm=TRUE)) %>%
  bind_rows(campus.pept) %>% mutate(percent_ept = round(pept,2))

f <- ggplot(eptt, aes(x=term_idx, y=percent_ept, col=term, group=term)) +
  geom_point() + geom_smooth(se=FALSE) + ylab("") + 
  scale_color_manual(values=c("black", term.col), name = "Term") + theme_bw() +
  ggtitle("Proportion of students per section who have completed\nthe EPT requirement prior to enrollment in CHEM") +
  xlab("Semester") +
  scale_y_continuous(labels = scales::percent_format(), limits = c(.8, 1))

ggplotly(f, tooltip = c("y", "x", "group"))

rm(f, eptt, campus.pept)
```
  

  
### College prep

    
```{r}
cpet <- chem %>%
  filter(term_idx != "NA") %>%
  group_by(term_idx, section) %>% 
  summarise(prop = mean(cp_engl, na.rm=TRUE)) %>%
  mutate(CP = "English")
cpmt <- chem %>%
  filter(term_idx != "NA") %>%
  group_by(term_idx, section) %>% 
  summarise(prop = mean(cp_math, na.rm=TRUE)) %>%
  mutate(CP = "Math")  
cplt <- chem %>%
  filter(term_idx != "NA") %>%
  group_by(term_idx, section) %>% 
  summarise(prop = mean(cp_lab, na.rm=TRUE)) %>%
  mutate(CP = "Lab")
cpt <- rbind(cpet, cpmt, cplt)

ggplot(cpt, aes(x=term_idx, y=prop, col=CP, group=CP)) + geom_point() +
  xlab("Semester") + ggtitle("Ave. # of college prep classes students have taken at admission per semester") +
  ylab("Average # of CP semesters") +
  geom_smooth(se=FALSE) + theme_bw() + scale_y_continuous()
```



```{r}
cont.varlist <- c('cp_engl', 'cp_math', 'cp_lab')

samp.means <- foreach(i = 1:3, .combine = rbind) %dopar% {
                mv.mean.wide(df=chem, var=cont.varlist[i], 
                             group='term_idx', col.lab=levels(chem$term_idx))
}

cp.out <- t(samp.means[c(2,4,6),])
colnames(cp.out) <- c("Engl", "Math", "Lab Sci")

kable(cp.out, align='ccc', caption = "Mean(SD) for number of college prep classes by term.")
```


    
## Academic Standing

### College
```{r}
rt <- chem %>%
  group_by(term_idx, college_name) %>%
  summarise(n=n()) %>%
  mutate(pct=n/sum(n))
ggplot(rt, aes(x=term_idx, y=pct, col=college_name, group=college_name)) + geom_point() +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0,.5)) + xlab("Semester") +
  ggtitle("Proportion of colleges students are in per semester") + ylab("% Colleges") +
  geom_smooth(se=FALSE) + theme_bw() + scale_color_discrete(name = "College Name")
```
  
  
### Student Level
```{r}
slt <- chem %>%
  filter(!is.na(stulevel)) %>%
  group_by(term_idx, stulevel) %>%
  summarise(n=n()) %>%
  mutate(percent=round(n/sum(n),2))

a <- ggplot(slt, aes(x=term_idx, y=percent, col=stulevel, group=stulevel)) + geom_point() +
      scale_y_continuous(labels = scales::percent_format(), limits = c(0,.75)) + xlab("Semester") +
      ggtitle("Proportion of students at each academic status per semester") + ylab("") +
      geom_smooth(se=FALSE) + theme_bw() + 
      scale_colour_discrete(name = "Student Level",
                            labels = c("Freshman", "Sophomore", "Junior", "Senior"))

ggplotly(a, tooltip = c("y", "x", "group"))

rm(a, slt)
```

  

### Units taken this semester

```{r}
utt <- chem %>% 
  filter(term_idx != "NA") %>%
  group_by(term, term_idx, section) %>% 
  summarise(mean = mean(term_units, na.rm=TRUE))
autt <- mean(chem$term_units, na.rm=TRUE)

ggplot(utt, aes(x=term_idx, y=mean, col=term, group=term)) +
  geom_point() + geom_smooth(se=FALSE) + ylab("Average units") + 
  scale_color_manual(values=term.col, name = "Term")+ theme_bw() +
  ggtitle("Average # of units students are taking while enroll in CHEM111 per section") +
  xlab("Semester") + 
  geom_hline(yintercept=autt) + scale_y_continuous()
```
  
  
```{r}
units_out <- biv.var(df=chem, var='term_units',  row.var='section', col.var='term_idx')
units_out1 <- units_out[1:3,1:8]
units_out2 <- units_out[c(1,2,4),c(1,9:15)]

kable(units_out1,  align='c', row.names=FALSE, caption='Mean(sd) student course load per section per term (09-12)')
kable(units_out2,  align='c', row.names=FALSE, caption='Mean(sd) student course load per section per term (13-16)')
```



